\bf Model Visualization Techniques for a Social Network Model

### 2017-08-24 paper.Rnw
add some transitions, add a table of model estimates

\bigskip
\begin{abstract}
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram. In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996). Unlike other network models, SAOMs are not very well understood. We use model visualization techniques introduced in Wickham et al (2015) in order to make them a little less murky. The SAOMs are a prime example of a set of models that can benefit greatly from application of model visualization. With the help of static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
\end{abstract}

@@

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.

XXX Need a transistion to next section XXX 
}

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statisticscomputed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
p1 <- plot(gof1, period = 1)
p2 <- plot(gof2, period = 1)
gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

@@

<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016. Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25. We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.height=3>>=
seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
ggplot(data = seobama) + 
  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party),size = 1) + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~senate, nrow = 1)
@

Because we require binary edges for our models, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as 
\[
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\]
The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}. 
}
\section{Model Visualizations} \label{sec:ModelVis}

@@

For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data. In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:

@@

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .5, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  #theme_net() + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, in the middle of her second term in the senate. In that time, she authored two bills and was a cosponsor on 17 other bills. With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram. We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator. We also map shape to sex, and keep color mapped to party of the senator. In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators. 
% ST: I don't like this paragraph
% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

@@

# make a df of wave 2, wave 1, and the three averages and facet. 
names(actual1)[1:2] <- c("from", "to")
names(actual2)[1:2] <- c("from", "to")
actual1$count <- 1
actual1$weight <- 1
actual2$count <- 1
actual2$weight <- 1
actual1 <- actual1 %>% dplyr::select(from,to, count, weight)
actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
actual1$cat <- "First Observation"
actual2$cat <- "Second Observation"

@@

p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
   labs(title = "Wave 1") + ThemeNet + 
    theme(plot.background = element_rect(color = 'black'))
p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
  labs(title = "Wave 2")  +  ThemeNet + 
  theme(plot.background = element_rect(color = 'black')) 
grid.arrange(p2, p1, p3, nrow = 1)
@

@@

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3>>=
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")

@@

<<getdata>>=
nulls <- read.csv("data/distribution_null_model.csv")
alt <- read.csv("data/distribution_jumpTT_model.csv")
alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
@

@@

simu_spread <- simu2 %>% spread(parameter, estimate)
ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + ThemeNoNet
@

@@

<<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side. There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=3>>=
#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(drink[20:35,])
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
data(M1ests_bigfriends)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
# M1sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M1eff,
#                        parms = as.numeric(M1parms), 
#                        N = 1000)
# M2sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M2eff,
#                        parms = as.numeric(M2parms), 
#                        N = 1000)
# M3sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M3eff,
#                        parms = as.numeric(M3parms), 
#                        N = 1000)
# M1simsdf <- sims_to_df(M1sims1000)
# M2simsdf <- sims_to_df(M2sims1000)
# M3simsdf <- sims_to_df(M3sims1000)
M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))

# M1avg$from <- paste0("V", M1avg$from)
# M1avg$to <- paste0("V", M1avg$to)
# M1avg2$from <- paste0("V", M1avg2$from)
# M1avg2$to <- paste0("V", M1avg2$to)
# M1avgW2$from <- paste0("V", M1avgW2$from)
# M1avgW2$to <- paste0("V", M1avgW2$to)
# M2avg$from <- paste0("V", M2avg$from)
# M2avg$to <- paste0("V", M2avg$to)
# M2avgW2$from <- paste0("V", M2avgW2$from)
# M2avgW2$to <- paste0("V", M2avgW2$to)
# M2avg2$from <- paste0("V", M2avg2$from)
# M2avg2$to <- paste0("V", M2avg2$to)
# M3avg$from <- paste0("V", M3avg$from)
# M3avg$to <- paste0("V", M3avg$to)
# M3avgW2$from <- paste0("V", M3avgW2$from)
# M3avgW2$to <- paste0("V", M3avgW2$to)
# M3avg2$from <- paste0("V", M3avg2$from)
# M3avg2$to <- paste0("V", M3avg2$to)
combinedavgs$linewidth[combinedavgs$cat %in% c("First Observation", "Second Observation")] <- .5
ggplot(data = combinedavgs) + 
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) + 
  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
  facet_wrap(~cat, nrow = 1)

>>>>

\bigskip
\begin{abstract}
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, ``The Small World Problem" by Stanley Milgram. In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996). Unlike other network models, SAOMs are not very well understood. We use model visualization techniques introduced in Wickham et al (2015) in order to make them a little less murky. The SAOMs are a prime example of a set of models that can benefit greatly from application of model visualization. With the help of static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
\end{abstract}

@@

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$. \st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function, or in our case the rate parameter, to fully describe the CTMC that is used to model network change.}
\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. Estimating the parameters in these models is difficult, but thanks to the SIENA software, we have an accessible way to fit SAOMs to network observations.

XXX HH can you help? I need a way better transition than that XXX
}

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
p1 <- plot(gof1, period = 1)
p2 <- plot(gof2, period = 1)
gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

@@

<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016. Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25. We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.height=3>>=
seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
ggplot(data = seobama) + 
  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party),size = 1) + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~senate, nrow = 1)
@

Because we require binary edges for our models, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as 
\[
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\]
The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}. In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these data sets use those models to guide our further exploration of SAOMs. 
}

\section{Model Visualizations} \label{sec:ModelVis}

@@

<<tableofmeans>>=
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
senM1est <- read.csv("data/congress/senateM1ests.csv")
senM2est <- read.csv("data/congress/senateM2ests.csv")
senM3est <- read.csv("data/congress/senateM3ests.csv")
senM1est$Model <- "M1"
senM2est$Model <- "M2"
senM3est$Model <- "M3"
senM1est$Sim <- 1:nrow(senM1est)
senM2est$Sim <- 1:nrow(senM2est)
senM3est$Sim <- 1:nrow(senM3est)
senM1est %>% gather(parameter, estimate, Rate:recip) %>% select(Sim, Model, parameter, estimate) -> senM1est2 
senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% select(Sim, Model, parameter, estimate) -> senM2est2 
senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% select(Sim, Model, parameter, estimate) -> senM3est2 
senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
senateEsts$parameter <- as.factor(senateEsts$parameter)
levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
SSE <-
  senateEsts %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
  mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(as.character(parameter))
SFE <- simu2 %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
   mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(parameter)
@

\begin{table}
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}

For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data. Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests} In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:

@@

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  #theme_net() + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term in the senate. In that time, she authored two bills and was a cosponsor on 17 other bills. With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram. We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator. We also map shape to gender, and keep color mapped to party of the senator. In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators. In this single visualization, we have viewed node information (number of bills authored by a senator and the gender of that senator), edge infortmation (the direction and strength of ties between two senators), and time (before and after Clinton left the senate). 
% ST: I don't like this paragraph
% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

@@

# make a df of wave 2, wave 1, and the three averages and facet. 
names(actual1)[1:2] <- c("from", "to")
names(actual2)[1:2] <- c("from", "to")
actual1$count <- 1
actual1$weight <- 1
actual2$count <- 1
actual2$weight <- 1
actual1 <- actual1 %>% dplyr::select(from,to, count, weight)
actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
actual1$cat <- "1st Wave"
actual2$cat <- "2nd Wave"

@@

p2 <- ggplot(data = filter(combinedavgs, cat == "1st Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
   labs(title = "Wave 1") + ThemeNet + 
    theme(plot.background = element_rect(color = 'black'))
p3 <- ggplot(data = filter(combinedavgs, cat == "2nd Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
  labs(title = "Wave 2")  +  ThemeNet + 
  theme(plot.background = element_rect(color = 'black')) 
grid.arrange(p2, p1, p3, nrow = 1)
@

@@

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3>>=
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")

@@

<<getdata>>=
nulls <- read.csv("data/distribution_null_model.csv")
alt <- read.csv("data/distribution_jumpTT_model.csv")
alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
@

@@

#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
simu_spread <- simu2 %>% spread(parameter, estimate)
ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.title.x = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0, inherit.blank = T),
                  axis.title.y = element_text(size = rel(1.5),
                                            face = 'plain', angle = 90, inherit.blank = T),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0, inherit.blank = T),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0, inherit.blank = T),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
@

@@

<<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side. There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(drink[20:35,])
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
data(M1ests_bigfriends)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
# M1sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M1eff,
#                        parms = as.numeric(M1parms), 
#                        N = 1000)
# M2sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M2eff,
#                        parms = as.numeric(M2parms), 
#                        N = 1000)
# M3sims1000 <- saom_simulate(dat = mydata, 
#                             struct = M3eff,
#                        parms = as.numeric(M3parms), 
#                        N = 1000)
# M1simsdf <- sims_to_df(M1sims1000)
# M2simsdf <- sims_to_df(M2sims1000)
# M3simsdf <- sims_to_df(M3sims1000)
M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))
M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
 group_by(from, to, wave) %>%
 summarise(count = n()) %>%
 mutate(weight = ifelse(from == to, 0, count))

# M1avg$from <- paste0("V", M1avg$from)
# M1avg$to <- paste0("V", M1avg$to)
# M1avg2$from <- paste0("V", M1avg2$from)
# M1avg2$to <- paste0("V", M1avg2$to)
# M1avgW2$from <- paste0("V", M1avgW2$from)
# M1avgW2$to <- paste0("V", M1avgW2$to)
# M2avg$from <- paste0("V", M2avg$from)
# M2avg$to <- paste0("V", M2avg$to)
# M2avgW2$from <- paste0("V", M2avgW2$from)
# M2avgW2$to <- paste0("V", M2avgW2$to)
# M2avg2$from <- paste0("V", M2avg2$from)
# M2avg2$to <- paste0("V", M2avg2$to)
# M3avg$from <- paste0("V", M3avg$from)
# M3avg$to <- paste0("V", M3avg$to)
# M3avgW2$from <- paste0("V", M3avgW2$from)
# M3avgW2$to <- paste0("V", M3avgW2$to)
# M3avg2$from <- paste0("V", M3avg2$from)
# M3avg2$to <- paste0("V", M3avg2$to)
combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
ggplot(data = combinedavgs) + 
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) + 
  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
  facet_wrap(~cat, nrow = 2)
